# More R poweR {#power}

## Objectives

- data import, writing
- for loops
- conditionals
- mention classes
- practice 
- what's next

You'll soon have questions that are outside the scope of this workshop, how do you find answers?


## Importing data

TODO

Remember you'll use `install.packages("package-name-in-quotes")` and then `library(package-name)`, and then you can explore the help or vignettes. And also, of course, Google to see how to use them!

- `readr` to read in .csv files
- `readxl` to read in Excel files
- `stringr` to work with strings
- `lubridate` to work with dates



## Subsetting data

TODO: change this from base-R

You will want to isolate bits of your data; maybe you want to just look at a single country or a few years. R calls this subsetting. There are several ways to do this. We'll go through a few options in base R so that you're familiar with them, and know how to read them. But then we'll move on to a new, better, intuitive, and game changing way with the `dplyr` package afterwards.

Remember your logical expressions from this morning? We'll use `==` here. 

### subsetting with base `[rows, columns]` notatation 
This notation is something you'll see a lot in base R. the brackets `[ ]` allow you to extract parts of an object. Within the brackets, the comma separates rows from columns. 
```{r, eval=FALSE}
## subset numeric data
gapminder[gapminder$lifeExp <32, ] # don't forget this comma! 

## subset factors
gapminder[gapminder$country == "Uruguay", ] # don't forget this comma! 
```
So our notation is saying 'select these rows, and all columns'.

We could also select which columns to keep using the `c()` function: 
```{r, eval=FALSE}
gapminder[gapminder$country == "Uruguay",
                     c("country", "year", "lifeExp")] # ?c: combines values into a vector or list
```

Contrast the above command with this one accomplishing the same thing:
```{r, eval=FALSE}
gapminder[1621:1632, ] # No idea what we are inspecting. Don't do this.

gapminder[1621:1632, c(1, 3, 4)] # Ditto. 
```

Yes, these both return the same result. But the second command is horrible for these reasons:

  * It contains [Magic Numbers](http://en.wikipedia.org/wiki/Magic_number_(programming)). The reason for keeping rows 1621 to 1632 will be non-obvious to someone else and that includes __you__ in a couple of weeks.
  * It is fragile. If the rows of `gapminder` are reordered or if some observations are eliminated, these rows may no longer correspond to the Uruguay data.

In contrast, the first command, is somewhat **self-documenting**; one does not need to be an R expert to take a pretty good guess at what's happening. It's also more robust. It will still produce the correct result even if `gapminder` has undergone some reasonable set of transformations (what if it were in in reverse alphabetical order?)

### subsetting with base `subset()` function
But we can improve how we subset by using the base `subset()` function, which can isolate pieces of an object for inspection or assignment. `subset()`'s main argument is also (unfortunately) called `subset`. 
```{r, eval=FALSE}
## subset gapminder
?subset
subset(gapminder, subset = country == "Mexico") # Ah, inspecting Mexico. Self documenting!
```
This returns all the columns. 

But what if you just want a few of the columns? You can also subset the columns you want. You can use `subset =` and `select =` together to simultaneously subset rows and columns or variables.
```{r, eval=FALSE}
subset(gapminder, subset = country == "Mexico", 
       select = c(country, year, lifeExp)) # ?c: combines values into a vector or list
```


## Repeating operations with for loops

Let's say we want to subset a few countries and plot pop through time. We could do it the way above, which would look like the following: 
```{r, eval=FALSE}
## plot population of some countries
mexico <- subset(gapminder, subset = country == "Mexico")
plot(mexico$year, mexico$pop)
dev.print(pdf, "mexico.pdf")

panama <- subset(gapminder, subset = country == "Panama") 
plot(panama$year, panama$pop)
dev.print(pdf, "panama.pdf")

ecuador <- subset(gapminder, subset = country == "Ecuador")
plot(ecuador$year, ecuador$pop)
dev.print(pdf, "ecuador")
```

But you can see already it's a lot of text, which means typo-prone and hard to read. Even if you copy-paste each one, there's a lot of copy-paste, and is very typo-prone. Plus, what if you wanted to instead plot lifeExp? You'd have to remember to change it each time...it gets messy quick. And we're just doing it with 3 countries here; what if we wanted to do it to all 142 countries? Eek. 


Better with a for loop. This will let us cycle through and do what we want to each thing in turn. If you want to iterate over a set of values, and perform the same operation on each, a `for` loop will do the job.

The basic structure of a `for` loop is:
```{r, eval=FALSE}
for(iterator in set of values){
  do a thing
}
```

Let's paste from what we had before, and modify it. Also, the `set of values` is the list of countries (`country_list`), and we want to iterate through each country (let's spell it `cntry` so it's distinctive). 
```{r, eval=FALSE}
for (cntry in country_list) {
  mexico <- subset(gapminder, subset = country == "Mexico") 
  plot(mexico$year, mexico$pop)
} 
```

We can't call it mexico anymore, but we could call it something more general. And let's comment the plot() line out while we build this, and add a print statement to see if it's behaving like we think it is. 
```{r, eval=FALSE}
for (cntry in country_list) {
  cntry_subset <- subset(gapminder, subset = country == cntry)  
  # plot(mexico$year, mexico$pop)
  print(cntry_subset)
} 
```

> Question: what is the variable `cntry_subset` right now, after running the for loop? 

Is this doing what we think it's doing? Let's create the country list and print the results each time to test our progress: 
```{r, eval=FALSE}
country_list <- c("Mexico", "Panama", "Ecuador") # identify the thing to loop through
for (cntry in country_list) {
  cntry_subset <- subset(gapminder, subset = country == cntry)  
  # plot(mexico$year, mexico$pop)
  print(cntry_subset)
} 
```

Excellent. Let's move on with the plot. 
```{r, eval=FALSE}
country_list <- c("Mexico", "Panama", "Ecuador") 
for (cntry in country_list) {
  cntry_subset <- subset(gapminder, subset = country == cntry) 
  plot(cntry_subset$year, cntry_subset$pop)
  dev.print(pdf, paste0(cntry,".pdf")) # ?paste0() will paste a string
} 
```

Great! And it doesn't matter if we just use these three countries or all the countries--let's try it. 

First let's create a figure directory and make sure it saves there since it's going to get out of hand quickly: 
```{r, eval=FALSE}
dir.create('figures') # this will be: software-carpentry/figures

country_list <- unique(gapminder$country) # ?unique() returns the unique values
for (cntry in country_list) {
  cntry_subset <- subset(gapminder, subset = country == cntry) 
  plot(cntry_subset$year, cntry_subset$pop)
  dev.print(pdf, paste0("figures/", cntry,".pdf")) # don't forget the `/`: it's a path!
} 
```

So that took a little longer than just the 3, but still super fast. For loops are sometimes just the thing you need to iterate over many things in your analyses. 

Now let's say we also want to record the mean population of each country. We'd add a line to the for loop, and comment out all the plotting for now (to save time, you could also just leave it): 
```{r, eval=FALSE}
dir.create('figures') # this will be: software-carpentry/figures

country_list <- unique(gapminder$country) 
for (cntry in country_list) {
  cntry_subset <- subset(gapminder, subset = country == cntry) 
  # plot(cntry_subset$year, cntry_subset$pop)
  # dev.print(pdf, paste0("figures/", cntry,".pdf"))
  
  pop_mean <- mean(cntry_subset$pop)
  print(paste('mean pop for', cntry, 'is', pop_mean))
} 
```

We know it worked since it printed correctly. But we didn't capture it: `cntry_subset` is just Zimbabwe. Let's create an object outside the loop and add to it each time. 
```{r, eval=FALSE}
dir.create('figures') # this will be: software-carpentry/figures

country_list <- unique(gapminder$country) # ?unique() returns the unique values
country_pop_mean <- data.frame()

for (cntry in country_list) {
  cntry_subset <- subset(gapminder, subset = country == cntry) 
  # plot(cntry_subset$year, cntry_subset$pop)
  # dev.print(pdf, paste0("figures/", cntry,".pdf")) 
  
  pop_mean <- mean(cntry_subset$pop)
  # print(paste('mean pop for', cntry, 'is', pop_mean))
  country_pop_mean <- rbind(country_pop_mean, data.frame(cntry, pop_mean))
} 
```
This approach can be useful, but 'growing your results' (building
the result object incrementally) is computationally inefficient, so avoid
it when you are iterating through a lot of values.

For loops can also lead to temporary variables that you don't need. But they can be really useful at times. 

## conditional statements with `if` and `else`

Often when we're coding we want to control the flow of our actions. This can be done
by setting actions to occur only if a condition or a set of conditions are met.
```{r, eval=FALSE}
# if
if (condition is true) {
  do something
}

# if ... else
if (condition is true) {
  do something
} else {  # that is, if the condition is false,
  do something different
}
```

Say, for example, that in addition to saving population figures for all countries, we want to save life expectancy figures for countries in Asia only. 
```{r, eval=FALSE}
dir.create('figures') # this will be: software-carpentry/figures

country_list <- unique(gapminder$country) # ?unique() returns the unique values
country_pop_mean <- data.frame()
for (cntry in country_list) {
  cntry_subset <- subset(gapminder, subset = country == cntry) 
  # plot(cntry_subset$year, cntry_subset$pop)
  # dev.print(pdf, paste0("figures/", cntry,".pdf")) 
  
  pop_mean <- mean(cntry_subset$pop)
  # print(paste('mean pop for', cntry, 'is', pop_mean))
  country_pop_mean <- rbind(country_pop_mean, data.frame(cntry, pop_mean))
  
  ## if Asia, calculate mean(lifeExp)
  if (unique(cntry_subset$continent) == "Asia") { # read: if (the continent is Asia) {then}
    plot(cntry_subset$year, cntry_subset$lifeExp) 
    dev.print(pdf, paste0("figures/", cntry, "_lifeExp.pdf")) # change the filename
  }
} 
```


And if the country is in Africa, let's plot the mean GDP. 
```{r, eval=FALSE}
dir.create('figures') # this will be: software-carpentry/figures

country_list <- unique(gapminder$country) # ?unique() returns the unique values
country_pop_mean <- data.frame()
for (cntry in country_list) {
  cntry_subset <- subset(gapminder, subset = country == cntry) 
  # plot(cntry_subset$year, cntry_subset$pop)
  # dev.print(pdf, paste0("figures/", cntry,".pdf")) 
  
  pop_mean <- mean(cntry_subset$pop)
  # print(paste('mean pop for', cntry, 'is', pop_mean))
  country_pop_mean <- rbind(country_pop_mean, data.frame(cntry, pop_mean))
  
  ## if Asia, calculate mean(lifeExp)
  if (unique(cntry_subset$continent) == "Asia") { # read: if (the continent is Asia) {then}
    plot(cntry_subset$year, cntry_subset$lifeExp) 
    dev.print(pdf, paste0("figures/", cntry, "_lifeExp.pdf")) 
  } else if (unique(cntry_subset$continent) == "Africa") {
    plot(cntry_subset$year, cntry_subset$gdpPercap) 
    dev.print(pdf, paste0("figures/", cntry, "_gdpPercap.pdf")) # change the filename
  }
} 
```
